Evolution of ORM Systems: Formal Foundations

### 2013-01-14 doc/description.tex
copyValue, copyRow

\subsection{Database Transformation}
The database is composed from two separated parts - database schema and data. Several transformation are defined to make manipulating a database easily.
\paragraph{Add Column} A new column can be add into a table if there are no name collisions. If the column has the $NOTNULL$ constraint, then the default value has to be added to all rows in the table.
\begin{align*}
	\mathbf{addColumn} : Label \times Column* \times Database \rightarrow Database \\ 	
	addColum(l, c, d) = \begin{cases}
		\inference{|c| > 1}{addColumn(l, head(c), d) \bullet addColumn(l, tail(c), d)} \\ \\
		\inference{|c| = 1 \wedge \exists t \in tables(d) \wedge name(t) = l \\ \wedge \forall f \in columns(t) : name(f) \neq name(c) \\ \wedge NOTNULL \not \in constraints(c) }{columns(t) \cup c}\\ \\
		\inference{|c| = 1 \wedge \exists t \in tables(d) \wedge name(t) = l \\ \wedge \forall f \in columns(t) : name(f) \neq name(c) \\ \wedge NOTNULL \in constraints(c) }{columns(t) \cup c}\\ \bullet appendValues(l, [name(c), defVal(c)]) \\ \\
		Inconsistent \;d 
	 \end{cases}\\ 
	 l \in Label, c \lhd Column*,d \lhd Database
\end{align*}

\paragraph{Copy Column} The transformation creates a copy of a column in a another table in a database. The transformation changes only the structure of database - stored data are not changed.
\begin{align*}
	\mathbf{copyColumn} : Label \times Label \times Label \times Database \rightarrow Database \\
	copyColumn(l_s, l_t, l_c, d) = \begin{cases}
 		\inference{\exists t \in tables(d) : name(t) = l_s \\ \wedge \exists c \in columns(t) : name(c) = l_c \\ \wedge \exists e \in tables(d) : name(e) = l_t  \\ \wedge \forall q \in properties(e) : name(q) \neq l_c}{columns(e) \cup c} 
	\end{cases} \\
	l_s,l_t,l_c \in Label \wedge l_s \neq l_t,d \lhd Database
\end{align*}
\paragraph{Add Table} The transformation creates a new table in the database. The new table should not contain any foreign keys.
\begin{align*}
	\mathbf{addTable} : Table* \times Database \rightarrow Database \\ 
	addTable(t, d) = \begin{cases}
		\inference{|t| > 1}{addTable(head(t),d) \bullet addTable(tail(t), d)}
\\ \\	
		\inference{|t| = 1 \wedge \forall e \in tables(d) \wedge name(t) \neq name(e) \\ \wedge foreignKeys(t) = \varnothing }{tables(d) \cup t}
\\ \\
		Inconsistent \;d 
	 \end{cases}\\ 
	t = Table*, d \lhd Database
\end{align*}
\paragraph{Insert row} The function allow to insert a row(s) into a database.
\begin{align*}
	\mathbf{insertRow} : Row* \times Database \rightarrow Database \\
	insertRow(r, d) = \begin{cases}
			\inference{|r|>1}{insertRow(head(r), d) \bullet insertRow(tail(r), d)} \\ \\
		 	\inference{|r|=1 \wedge \exists t \in tables(d) : name(t) = reference(r) \\ \wedge \forall e \in tableData(t, d) : pkValue(e) \neq pkValue(r)}{data(d) \cup r} \\ \\
			 Inconsistent \; d
		 \end{cases}	\\
		 r = Row*,d \lhd Database
\end{align*}
\paragraph{Append Values} The function appends a pair to all rows of a table.
\begin{align*}
	\mathbf{appendValues} : Label \times Pair \times Database \rightarrow Database \\
	appemdValues(l, p, d) = \begin{cases}
 		\inference{\exists t \in tables(d) : name(t) = l, \\ r = tableData(l, d)}{insertValue(r, p, d)} \\ \\
 		Inconsistent \; d
 	\end{cases}\\ \\
 	l \in Label, d \lhd Database
\end{align*}
\paragraph{Insert Value} The function inserts a new pair into an existing row.
\begin{align*}
	\mathbf{insertValues} : Row \times Pair* \times Database \rightarrow Daatabase \\
	insertValues(r, p, d) = \begin{cases}
 		\inference{|p| > 1}{insertValues(r, head(p), d)\bullet insertValues(r, tail(p), d)}  \\ \\
 		\inference{|p| = 1 \wedge \forall e \in pairs(r) : reference(e) \neq reference(p) \\ \wedge \exists t \in tables(d) : name(t) = refernces(r), \\ \wedge \exists c \in columns(t) : reference(p) = name(c)}{pairs(r) \cup p} \\ \\
	 	 Inconsistent \; d
 		\end{cases} \\
r \lhd Row, p = Pair*,d \lhd Databases
\end{align*}
\paragraph{COPY VALUES}
\begin{align*}
	\mathbf{copyValue} : Label \times Label \times Label \times (Instancece \times Instance) \times Database \times Database \\
	copyValue(l_s, l_t, l_{sc}, l_{tc} f_m, d) = \begin{cases}
 		\inference{\exists t_s \in tables(d) : name(t_s) = l_s, \\ t_t \in tables(d) : name(t_t) = l_t, \\ c \in columns(t_s) : name(c) = l_{sc}, \\ e \in columns(t_t) : name(e) = l_{tc} }{\begin{gathered}
			\forall p \in f_m(t_s, t_t)  : \\ insertValues(name(second(p)), \\ [l_{tc},value(pairByRef(l_{sc}, first(p))),d)
		\end{gathered}}
 	\end{cases}
\end{align*}

@@

Whereas the $insertProp$ on database level can change stored data too, because a mandatory property can be added into a class with already existing instances, then a default value has to be added to all existing instances.
\begin{align*}
insertProp(p, d) = \begin{cases}
\inference{!\:mandatory(p) \vee !\:instantiated(owningClass(p))}{addColumn(l, \rho(p,d), d)} 
\\ \\ 
\inference{ \wedge mandatory(p) \vee instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered} addColumn(l, \rho(p,d), d) \\ \bullet \forall r \in data(d), reference(r) = name(t) : \\ appendValue(name(t), [name(p), defaultValue(p)], d)
\end{gathered}
} 
\\ \\
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge mandatory(p) \wedge instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered}
addTable(\rho(p,d), d) \bullet \forall r \in row(owningClass(p)) :\\ insertRow(row(name(p), ["value", defaultValue(p)], \\ [name(p), value(primaryKey(r))]),d) 
\end{gathered}
} 
\end{cases}\\
p \in Properties, d \in Dataabses
\end{align*}

@@

\subsection{Software Alternation}
\paragraph{Rename Class}
\begin{align*}
RenameClass : Label \times Label \times State \rightarrow State \\ 
\Psi(RenameClass(l_o, l_n), s) = application(s) \circ renClass(l_o, l_n) \\
\Phi(RenameClass(l_o, l_n), s) = database(s) \bullet alterClass(l_o, l_n) \\ \\
\end{align*}

\begin{align*}
alterTable:  Label \times Label \times Database \rightarrow Database \\
alterTable(l_o, l_n, d) = \inference{\exists t \in tables(d) : name(d) = l_o \\ \wedge \forall e \in tables(d) : name(e) \neq l_n}{\begin{gathered}
t.name = l_n \\ \wedge \forall e \in tables(a) \wedge \forall (f \in fks(e) \wedge reference(f) = l_o ) \\ : f.reference = l_n 
\end{gathered}}\\
l_o, l_n \in Label, s \in States, a \lhd Applications, d \lhd Database
\end{align*}

\paragraph{Rename Property}
\begin{align*}
	RenameProperty : Label \times Label \times Label \times State \rightarrow State \\ \\
\Psi(RenameProperty(l_c, l_o, l_n, s)) = application(s) \circ renProp(l_c, l_o, l_n) \\
\Phi(RenameProperty(l_c, l_o, l_n, s)) = database(s) \bullet alterColumn(l_c, l_o, l_n) \\\\
\end{align*}

\begin{align*}
alterColumn : Label \times Label \times Label \times Database \rightarrow Database \\
	alterColumn(l_c, l_o, l_n, d) = \begin{cases}
 \inference{\exists t \in tables(a) : name(t) = l_c \\ \wedge \exists c \in columns(t) : name(c) = l_o \\ \wedge \forall e \in columns(t) : name(e) \neq l_n}{\begin{gathered}
c.name = l_n 
\end{gathered}}
\\ \\
\inference{\exists t \in tables(a) : name(t) = l_o \\ \wedge \forall e \in tables(t) : name(e) \neq l_n}{\begin{gathered}
t.name = l_n 
\end{gathered}}
 \end{cases}
 \\
l_c, l_o, l_n \in Label, a \lhd Applications, s \in States, d \lhd Database
\end{align*}

\paragraph{Rename Association}
\begin{align*}
RenameAssociation : Label \times Label \times Label \times State \rightarrow State \\
\Psi(RenameAssocaiation(l_c, l_o, l_n), s) = application(s) \circ renAssoc(l_c, l_o, l_n)\\
\Phi(RenameAssocaiation(l_c, l_o, l_n), s) = \begin{cases}
	\inference{\mathbf{TODO}}{database(s) \bullet alterTable(l_c, l_o, l_n)}
 \\
  \inference{\mathbf{TODO}}{database(s) \bullet alterColumn(l_c, l_o, l_n)}
 \end{cases}
\\\\
alterColumn : Label \times Label \times Label \times Database \rightarrow Database \\
%	alterAssoc(l_c, l_o, l_n, d) = \begin{cases}
%?? \inference{\exists t \in tables(a) : name(t) = l_c \\ \wedge \exists f \in fks(t) : name(c) = l_o \\ \wedge \forall e \in fks(t) : name(e) \neq l_n}{\begin{gathered}
%f.name = l_n 
%\end{gathered}}
%\\ \\
%\inference{\exists t \in tables(a) : name(t) = l_o \\ \wedge \forall e \in tables(t) : name(e) \neq l_n}{\begin{gathered}
%t.name = l_n 
%\end{gathered}}
% \end{cases}
% \\
l_c, l_o, l_n \in Label, a \lhd Applications, s \in Sates, d \lhd Database
\end{align*}

@@

>>>>

\subsection{Database Transformation}
The database is composed from two separated parts - database schema and data. Transformations are defined to manipulate the database schema and to manipulate stored data. These transformation were created with respect to evolutionary transformation, thus they ....

\paragraph{Add Column} A new column can be add into a table if there are no name collisions. If the column has the $NOTNULL$ constraint, then the default value has to be added to all rows in the table.
\begin{align*}
	\mathbf{addColumn} : Label \times Column* \times Database \rightarrow Database \\ 	
	addColum(l, c, d) = \begin{cases}
		\inference{|c| > 1}{addColumn(l, head(c), d) \bullet addColumn(l, tail(c), d)} \\ \\
		\inference{|c| = 1 \wedge \exists t \in tables(d) \wedge name(t) = l \\ \wedge \forall f \in columns(t) : name(f) \neq name(c) \\ \wedge NOTNULL \not \in constraints(c) }{columns(t) \cup c}\\ \\
		\inference{|c| = 1 \wedge \exists t \in tables(d) \wedge name(t) = l \\ \wedge \forall f \in columns(t) : name(f) \neq name(c) \\ \wedge NOTNULL \in constraints(c) }{columns(t) \cup c}\\ \bullet addValueToTable(l, [name(c), defVal(c)]) \\ \\
		Inconsistent \;d 
	 \end{cases}\\ 
	 l \in Label, c \lhd Column*,d \lhd Database
\end{align*}

\paragraph{Copy Column} The transformation creates a copy of a column in a another table in a database. The transformation changes only the structure of database - stored data are not changed.
\begin{align*}
	\mathbf{copyColumn} : Label \times Label \times Label \times Database \rightarrow Database \\
	copyColumn(l_s, l_t, l_c, d) = \begin{cases}
 		\inference{\exists t \in tables(d) : name(t) = l_s \\ \wedge \exists c \in columns(t) : name(c) = l_c \\ \wedge \exists e \in tables(d) : name(e) = l_t  \\ \wedge \forall q \in properties(e) : name(q) \neq l_c}{columns(e) \cup c} 
	\end{cases} \\
	l_s,l_t,l_c \in Label \wedge l_s \neq l_t,d \lhd Database
\end{align*}

\paragraph{Alter Column Name}
\begin{align*}
\mathbf{alterColumn} : Label \times Label \times Label \times Database \rightarrow Database \\
	alterColumn(l_c, l_o, l_n, d) =
 \inference{\exists t \in tables(a) : name(t) = l_c \\ \wedge \exists c \in columns(t) : name(c) = l_o \\ \wedge \forall e \in columns(t) : name(e) \neq l_n}{\begin{gathered}
c.name = l_n 
\end{gathered}}
 \\
l_c, l_o, l_n \in Label, d \lhd Database
\end{align*}

\paragraph{Add Table} The transformation creates a new table in the database. The new table should not contain any foreign keys.
\begin{align*}
	\mathbf{addTable} : Table* \times Database \rightarrow Database \\ 
	addTable(t, d) = \begin{cases}
		\inference{|t| > 1}{addTable(head(t),d) \bullet addTable(tail(t), d)}
\\ \\	
		\inference{|t| = 1 \wedge \forall e \in tables(d) \wedge name(t) \neq name(e) \\ \wedge foreignKeys(t) = \varnothing }{tables(d) \cup t}
\\ \\
		Inconsistent \;d 
	 \end{cases}\\ 
	t = Table*, d \lhd Database
\end{align*}
\paragraph{Insert row} The function allow to insert a row(s) into a database.
\begin{align*}
	\mathbf{insertRow} : Row* \times Database \rightarrow Database \\
	insertRow(r, d) = \begin{cases}
			\inference{|r|>1}{insertRow(head(r), d) \bullet insertRow(tail(r), d)} \\ \\
		 	\inference{|r|=1 \wedge \exists t \in tables(d) : name(t) = reference(r) \\ \wedge \forall e \in tableData(t, d) : pkValue(e) \neq pkValue(r)}{data(d) \cup r} \\ \\
			 Inconsistent \; d
		 \end{cases}	\\
		 r = Row*,d \lhd Database
\end{align*}
\paragraph{Add Value To Table} The function appends a pair to all rows of a table.
\begin{align*}
	\mathbf{addValueToTable} : Label \times Pair \times Database \rightarrow Database \\
	addValueToTable(l, p, d) = \begin{cases}
 		\inference{\exists t \in tables(d) : name(t) = l}	{\begin{gathered}
		insertValue(head(tableData(l, d)), p, d) \\ \bullet addValueToTable(tail(tableData(l, d)), p, d)
	\end{gathered}} \\ \\
 		Inconsistent \; d
 	\end{cases}\\ \\
 	l \in Label, p \lhd Pair,  d \lhd Database
\end{align*}

\paragraph{Insert Value} The function inserts a new pair into an existing row.
\begin{align*}
	\mathbf{insertValues} : Row \times Pair* \times Database \rightarrow Daatabase \\
	insertValues(r, p, d) = \begin{cases}
 		\inference{|p| > 1}{insertValues(r, head(p), d)\bullet insertValues(r, tail(p), d)}  \\ \\
 		\inference{|p| = 1 \wedge \forall e \in pairs(r) : reference(e) \neq reference(p) \\ \wedge \exists t \in tables(d) : name(t) = refernces(r), \\ \wedge \exists c \in columns(t) : reference(p) = name(c)}{pairs(r) \cup p} \\ \\
	 	 Inconsistent \; d
 		\end{cases} \\
r \lhd Row, p = Pair*,d \lhd Databases
\end{align*}
\paragraph{Copy Value} The function copy values from one column in a table to another column in the same or in another table - according to mapping function $f_m$. 
\begin{align*}
	\mathbf{copyValue} : Label \times Label \times Label \times (Row \times Row) \times Database \rightarrow Database \\
	copyValue(l_s, l_t, l_{sc}, l_{tc} f_m, d) = \begin{cases}
		\inference{\exists t_s \in tables(d) : name(t_s) = l_s, \\ t_t \in tables(d) : name(t_t) = l_t, \\ c \in columns(t_s) : name(c) = l_{sc}, \\ e \in columns(t_t) : name(e) = l_{tc}}{\begin{gathered}
			insertValues(second(p), \\ [l_{tc},value(pairByRef(l_{sc}, first(p)))],d)
		\end{gathered}}
 	\end{cases}
\end{align*}

\paragraph{Copy rows}
\begin{align*}
	\mathbf{copyRows} : Label \times Label \times (Row \times Row) \times Database \rightarrow Database \\
	copyRows(l_s, l_t, f_m, d) = \begin{cases}
 	\inference{|f_m| = \varnothing}{d} \\\\
 	\inference{|f_m| = 1 \\ \wedge \exists t_s, t_t \in tables(d) : name(t_s) = l_s, name(t_t) = l_t \\ \wedge t_s \cong t_t }{data(d) \cup insertRow(changeRef(l_t, first(f_m)), d)} \\\\
 	\inference{|f_m| > 1}{\begin{gathered}
		 d \bullet copyRows(l_s, l_t, head(f_m)) \bullet copyRows(l_s, l_t, tail(f_m))
	\end{gathered}}
  \end{cases}
\end{align*}

%\begin{align*}
%	v(l, fm, d) = \begin{cases}
%	\inference{fm = \varnothing}{d}\\ \\
%	\inference{|fm| > 1}{v(l, head(fm)) \bullet v(l, tail(fm))}\\ \\
%	\inference{|fm| = 1 \exists c \in pairs(first(fm)) : reference(c) = l \\ \wedge \exists e \in pairs(second(fm)) : reference(e) = l}{e \cup pairByRef(l,c)} 
% \end{cases}
%\end{align*}

@@

Whereas the $insertProp$ on database level can change stored data too, because a mandatory property can be added into a class with already existing instances, then a default value has to be added to all existing instances.
\begin{align*}
insertProp(p, d) = \begin{cases}
\inference{!\:mandatory(p) \vee !\:instantiated(owningClass(p))}{addColumn(l, \rho(p,d), d)} 
\\ \\ 
\inference{ \wedge mandatory(p) \vee instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered} addColumn(l, \rho(p,d), d) \\ \bullet \forall r \in data(d), reference(r) = name(t) : \\ addValueToTable(name(t), [name(p), defaultValue(p)], d)
\end{gathered}
} 
\\ \\
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge mandatory(p) \wedge instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered}
addTable(\rho(p,d), d) \bullet \forall r \in row(owningClass(p)) :\\ insertRow(row(name(p), ["value", defaultValue(p)], \\ [name(p), value(primaryKey(r))]),d) 
\end{gathered}
} 
\end{cases}\\
p \in Properties, d \in Dataabses
\end{align*}

@@

\subsection{Software Alternation}
\paragraph{Rename Class}
\begin{align*}
RenameClass : Label \times Label \times State \rightarrow State \\ 
\Psi(RenameClass(l_o, l_n), s) = application(s) \circ renClass(l_o, l_n) \\
\Phi(RenameClass(l_o, l_n), s) = database(s) \bullet alterClass(l_o, l_n) \\ \\
\end{align*}
\begin{align*}
alterTable:  Label \times Label \times Database \rightarrow Database \\
alterTable(l_o, l_n, d) = \inference{\exists t \in tables(d) : name(d) = l_o \\ \wedge \forall e \in tables(d) : name(e) \neq l_n}{\begin{gathered}
t.name = l_n \\ \wedge \forall e \in tables(a) \wedge \forall (f \in fks(e) \wedge reference(f) = l_o ) \\ : f.reference = l_n 
\end{gathered}}\\
l_o, l_n \in Label, s \in States, a \lhd Applications, d \lhd Database
\end{align*}

\paragraph{Rename Property}
\begin{align*}
	RenameProperty : Label \times Label \times Label \times State \rightarrow State \\ \\
\Psi(RenameProperty(l_c, l_o, l_n, s)) = application(s) \circ renProp(l_c, l_o, l_n) \\
\Phi(RenameProperty(l_c, l_o, l_n, s)) = \begin{cases}
 	\inference{cardinality(propertyByName(l_c, l_o, application(s))) \leq 1}{database(s) \bullet alterColumn(l_c, l_o, l_n))} \\\\
 	\inference{cardinality(propertyByName(l_c, l_o, application(s))) > 1}{alterTable(l_o, l_n, database(s))}
 \end{cases}\\
l_c, l_o, l_n \in Label,  s \lhd Software
\end{align*}

\paragraph{Rename Association}
\begin{align*}
RenameAssociation : Label \times Label \times Label \times State \rightarrow State \\
\Psi(RenameAssocaiation(l_c, l_o, l_n), s) = application(s) \circ renAssoc(l_c, l_o, l_n)\\
\Phi(RenameAssocaiation(l_c, l_o, l_n), s) = \begin{cases}
	\inference{\exists a  \in getAssocByname(application(s)) : \\
  name(a) = l_c \\ \wedge startCardinality(a) > 1 }{database(s) \bullet alterTable(l_c, l_o, l_n)}
 \\ \\
  \inference{\exists a  \in getAssocByname(application(s)) : \\
  name(a) = l_c \\ \wedge startCardinality(a) \leq 1 }{\begin{gathered}
database(s) \\ \bullet alterForeignKey(endClass(a) ,l_c, l_o, l_n)
\end{gathered}}
 \end{cases}
\\\\
%	alterAssoc(l_c, l_o, l_n, d) = \begin{cases}
%?? \inference{\exists t \in tables(a) : name(t) = l_c \\ \wedge \exists f \in fks(t) : name(c) = l_o \\ \wedge \forall e \in fks(t) : name(e) \neq l_n}{\begin{gathered}
%f.name = l_n 
%\end{gathered}}
%\\ \\
%\inference{\exists t \in tables(a) : name(t) = l_o \\ \wedge \forall e \in tables(t) : name(e) \neq l_n}{\begin{gathered}
%t.name = l_n 
%\end{gathered}}
% \end{cases}
% \\
l_c, l_o, l_n \in Label, a \lhd Applications, s \in Sates, d \lhd Database
\end{align*}

@@

\section{Refactoring Cases}
\subsection{Move Property}
\subsection{Extract and Inline Class}
